<h2>Some methods for the comparative analysis of discrete character data</h2>

```{r, dpi=300}
library(phytools)
packageVersion("phytools") ## installed from GitHub 0.6-99 is fine
```

<h3><i>Method 1</i>: Testing for heterogeneous rates of discrete character 
evolution on phylogenies.</h3>

In this first of three short modules, we'll see how to fit a extended 
M<i>k</i> model in which the rate of character evolution varies as a 
function of 'regimes' mapped onto the tree.

We can do this first using some simulated data that can be downloaded here:

1. <a href="simulated-tree.tre">simulated-tree.tre</a>
2. <a href="simulated-data.csv">simulated-data.csv</a>

```{r, dpi=300}
sim.tree<-read.tree("simulated-tree.tre")
sim.data<-read.csv("simulated-data.csv",row.names=1)
```

Plot our tree:

```{r, dpi=300}
plotTree(sim.tree,ftype="off")
nodelabels(cex=0.6)
```

Now let's imagine that we had some <i>a priori</i> reason, independent of our
character data, to suspect that clade descended from node '93' had evolved 
under a different evolution. Let's map this hypothesis on the tree, jointly
with our trait data at the tips:

```{r, dpi=300}
hypothesis<-paintSubTree(sim.tree,93,
	"regime 2","regime 1")
cols<-setNames(c("blue","red"),c("regime 1","regime 2"))
plot(hypothesis,cols,ftype="off")
trait<-setNames(sim.data[,1],rownames(sim.data))
tiplabels(pie=to.matrix(trait[hypothesis$tip.label],
	levels(trait)),cex=0.3,
	piecol=c("white","grey","black"))
legend("topleft",pch=21,pt.cex=2,
	pt.bg=c("white","grey","black"),
	legend=levels(trait),bty="n")
```

Let's fit two models to the data as follows:

```{r, dpi=300}
singleRateModel<-fitMk(hypothesis,trait,
	model="ER")
singleRateModel
multiRateModel<-fitmultiMk(hypothesis,trait,
	model="ER")
multiRateModel
```

These models can be compared directly:

```{r, dpi=300}
data.frame(model=c("single-rate","multi-rate"),
	logLik=c(logLik(singleRateModel),
	logLik(multiRateModel)),
	k=c(attr(AIC(singleRateModel),"df"),
	attr(AIC(multiRateModel),"df")),
	AIC=c(AIC(singleRateModel),
	AIC(multiRateModel)))
```

This tells us that the multi-rate model is much better supported than the
single rate model.

Let's proceed to plot our best-fitting model as follows:

```{r, dpi=300}
layout(matrix(c(1,1,2,3),2,2,byrow=TRUE))
plot(hypothesis,cols,ftype="off",
	direction="downwards")
tiplabels(pie=to.matrix(trait[hypothesis$tip.label],
	levels(trait)),cex=0.3,
	piecol=c("white","grey","black"))
legend("topleft",pch=21,pt.cex=2,
	pt.bg=c("white","grey","black"),
	legend=levels(trait),bty="n")
obj<-multiRateModel
obj$rates<-round(obj$rates[1],2)
obj$regimes<-NULL
class(obj)<-"fitMk"
plot(obj,show.zeros=FALSE,
	mar=rep(2.1,4),show.zeros=FALSE,
	tol=1e-3,cex.traits=0.9,
	cex.rates=0.6)
mtext(text="a) Regime 1 (blue)",
    adj=0,line=-1,cex=0.9)

obj<-multiRateModel
obj$rates<-round(obj$rates[2],2)
obj$regimes<-NULL
class(obj)<-"fitMk"
plot(obj,show.zeros=FALSE,
	mar=rep(2.1,4),show.zeros=FALSE,
	tol=1e-3,cex.traits=0.9,
	cex.rates=0.6)
mtext(text="b) Regime 2 (red)",
    adj=0,line=-1,cex=0.9)
```

Now we can try the same thing with a real data set consisting of a phylogeny
of <i>Anolis</i> lizards and  the number of vertebrae in the tail:

1. <a href="VERT.CSV">VERT.CSV</a>
2. <a href="ANOLIS.PHY">ANOLIS.PHY</a>

Start by reading the data from file:

```{r, dpi=300}
X<-read.csv("VERT.CSV",row.names=1)
vert<-factor(setNames(X[,1],rownames(X)),
	levels=min(X[,1]):max(X[,1]))
```

Now we can create the model that we want to fit. For this step, I propose
an ordered reversible model - in which caudal vertebrae are gained & lost 
sequentially - although possibly with different rates in either direction:

```{r, dpi=300}
k<-length(levels(vert))
ordered<-matrix(0,k,k,dimnames=list(levels(vert),
	levels(vert)))
for(i in 1:k){
	if(i<k) ordered[i,i+1]<-1
	if(i>1) ordered[i,i-1]<-2
}
ordered[i,i-1]<-2
```

This is the design matrix of our model:

```{r, dpi=300}
ordered
```

Now let's convert our vector of vertebrae to a matrix. We do this because
there are certain levels of our trait that are not present in the input
data vector:

```{r, dpi=300}
vert<-to.matrix(vert,levels(vert))
head(vert)
```

Our next step is to propose our hypothesis for rate variation on the tree.
In this case, to keep it simple, we propose that mainland & Caribbean
anoles have different rates. I have saved this hypothesis (mapped on the 
phylogeny) to our tree file:

```{r, dpi=300}
anolis.tree<-read.simmap("ANOLIS.PHY",format="phylip")
plot(anolis.tree,ftype="off",type="fan",
	colors=setNames(c("blue","brown"),
	c("I","M")))
```

Unfortunately, the taxa in our tree & our data matrix don't match exactly.
We can identify mismatches using <code>geiger::name.check</code> & fix 
them as follows:

```{r, dpi=300}
library(geiger)
chk<-name.check(anolis.tree,vert)
anolis.tree<-drop.tip.simmap(anolis.tree,
	chk$tree_not_data)
vert<-vert[anolis.tree$tip.label,]
name.check(anolis.tree,vert)
```

Now we should be ready to fit our two models:

```{r, dpi=300}
fit.single<-fitMk(anolis.tree,vert,model=ordered)
fit.multi<-fitmultiMk(anolis.tree,vert,model=ordered)
```

Here's a summary of our fitted models:

```{r, dpi=300}
data.frame(model=c("single-rate","multi-rate"),
	logLik=c(logLik(fit.single),
	logLik(fit.multi)),
	k=c(attr(AIC(fit.single),"df"),
	attr(AIC(fit.multi),"df")),
	AIC=c(AIC(fit.single),
	AIC(fit.multi)))
```

Here, the two-rate model <i>is not</i> justified. Let's nonetheless
graph our fitted model as we did in the previous part of the exercise:

```{r, dpi=300, fig.height=12,fig.width=7}
par(mfrow=c(2,1))
obj<-fit.multi
obj$rates<-round(obj$rates[3:4],2)
obj$regimes<-NULL
class(obj)<-"fitMk"
plot(obj,show.zeros=F,mar=rep(2.1,4),show.zeros=F,
	tol=1e-3,cex.traits=0.8,cex.rates=0.4)
mtext(text="a) Mainland caudal vertebra number",
    adj=0,line=-1,cex=0.9)
obj<-fit.multi
obj$rates<-round(obj$rates[1:2],2)
obj$regimes<-NULL
class(obj)<-"fitMk"
plot(obj,show.zeros=F,mar=rep(2.1,4),show.zeros=F,
	tol=1e-3,cex.traits=0.8,cex.rates=0.4)
mtext(text="b) Caribbean caudal vertebra number",
    adj=0,line=-1,cex=0.9)
```

<h3><i>Method 2</i>: Fitting discrete character evolution to data with 
intraspecific polymorphism.</h3>

For the second module we can use a new function called 
<code>fitpolyMk</code>. This function handles discrete character data
with intraspecific polymorphism in a fairly obvious way - that is merely as 
an intermediate state between the two or more observed character values.

The data for this part of the exercise can be downloaded here:

1. <a href="polymorphic-data.csv">polymorphic-data.csv</a>
2. <a href="polymorphic-tree.phy">polymorphic-tree.phy</a>

```{r, dpi=300}
poly.tree<-read.tree("polymorphic-tree.phy")
poly.data<-read.csv(file="polymorphic-data.csv",row.names=1)
y<-setNames(poly.data[,1],rownames(poly.data))
```

These are what our data should look like:

```{r, dpi=300}
y
```

Now let's plot them:

```{r, dpi=300}
plotTree(poly.tree,ftype="off",lwd=1,type="fan")
X<-strsplit(setNames(as.character(y),names(y)),"+",fixed=TRUE)
pies<-matrix(0,Ntip(poly.tree),4,dimnames=list(poly.tree$tip.label,
	c("A","B","C","D")))
for(i in 1:Ntip(poly.tree)) 
	pies[poly.tree$tip.label[i],X[[poly.tree$tip.label[i]]]]<-
		rep(1/length(X[[poly.tree$tip.label[i]]]),
		length(X[[poly.tree$tip.label[i]]]))
tiplabels(pie=pies,piecol=c("black","yellow","red","blue"),
	cex=0.35)
legend(x="topleft",legend=c("A","B","C","D"),pt.cex=2,pch=21,
	pt.bg=c("black","yellow","red","blue"))
```

Our next step is just to fit our different models. We can plot each of
these as we go:

```{r, dpi=300}
par(mfrow=c(2,2),fg="blue")
er.ordered<-fitpolyMk(poly.tree,y,model="ER",ordered=TRUE)
plot(er.ordered,lwd=2,mar=c(1.1,1.1,2.1,1.1))
mtext("a) fitted ER \'ordered\' model",adj=0,
	line=-0.5,col="black")
er.unordered<-fitpolyMk(poly.tree,y,model="ER")
plot(er.unordered,lwd=2,mar=c(1.1,1.1,2.1,1.1))
mtext("b) fitted ER \'unordered\' model",adj=0,
	line=-0.5,col="black")
transient.ordered<-fitpolyMk(poly.tree,y,model="transient",
	ordered=TRUE)
plot(transient.ordered,lwd=2,mar=c(1.1,1.1,2.1,1.1))
mtext("c) fitted transient \'ordered\' model",adj=0,
	line=-0.5,col="black")
transient.unordered<-fitpolyMk(poly.tree,y,
	model="transient")
plot(transient.unordered,lwd=2,mar=c(1.1,1.1,2.1,1.1))
mtext("d) fitted ER \'unordered\' model",adj=0,
	line=-0.5,col="black")
par(fg="black")
```

Compare our different models:

```{r, dpi=300}
data.frame(transition_model=c("ER","ER","transient","transient"),
	ordered=c("no","yes","no","yes"),
	logLik=c(logLik(er.unordered),logLik(er.ordered),
	logLik(transient.unordered),logLik(transient.ordered)),
	k=c(attr(AIC(er.unordered),"df"),attr(AIC(er.ordered),"df"),
	attr(AIC(transient.unordered),"df"),attr(AIC(transient.ordered),"df")),
	AIC=c(AIC(er.unordered),AIC(er.ordered),
	AIC(transient.unordered),AIC(transient.ordered)))
```

This tells us that our equal-rates ordered model is the best fit to our
data.

<h3><i>Method 3</i>: Modeling discrete character evolution using Bayesian 
MCMC in R.</h3>

In this final short module we can use the <i>phytools</i> Bayesian MCMC function 
for the extended M<i>k</i> model which is called <code>mcmcMk</code>. Note that
this is also possible to do using the <i>diversitree</i> package.

In this case we'll use the following tree & data:

1. <a href="elopomorph.tre">elopomorph.tre</a>
2. <a href="elopomorph.csv">elopomorph.csv</a>

```{r, dpi=300}
X<-read.csv("elopomorph.csv",row.names=1)
feed.mode<-setNames(X[,1],rownames(X))
feed.mode
eel.tree<-read.tree("elopomorph.tre")
eel.tree
mcmc<-mcmcMk(eel.tree,feed.mode,model="ARD",
	prior.rate=100,prop.var=0.001,ngen=10000)
```

Now let's plot our results:

```{r, dpi=300}
par(mfrow=c(2,2))
mar<-c(4.1,4.1,2.1,1.1)

plot.new()
par(mar=mar)
plot.window(xlim=c(0,1),ylim=c(0,1))
library(png)
download.file(
	"http://www.phytools.org/evol2019/Enchelycore_schismatorhynchus.png",
	"eel-picture.png",mode="wb")
img<-readPNG(source="eel-picture.png")
rasterImage(img,0,0,1,1)
mtext(text="a) a biting eel",adj=0,line=0,cex=1)

plotTree(eel.tree,fsize=0.5,ftype="i",
	ylim=c(-8,Ntip(eel.tree)),mar=mar)
FMODE<-to.matrix(feed.mode,levels(feed.mode))
par(fg="transparent")
tiplabels(pie=FMODE[eel.tree$tip.label,],piecol=c("red","blue"),cex=0.3)
par(fg="black")
par(cex=0.8)
add.simmap.legend(colors=setNames(c("red","blue"),c(" bite"," suction")),
	vertical=FALSE,fsize=0.6,prompt=FALSE,x=2,y=-8)
mtext(text="b) phylogeny of eels",
	adj=0,line=0,cex=1)

par(mar=mar)
plot(mcmc,main="")
mtext(text="c) likelihood profile from MCMC",adj=0,
	line=1,cex=1)

d<-density(mcmc)
plot(d,main="")
mtext(text=expression(paste(
	"d) estimated posterior density for ",Q[ij])),
	adj=0,line=1,cex=1)
```

That's it.

